---
title: "01a: Wrangling & plotting"
subtitle: "Bayesian regression: theory & practice"
author: "Michael Franke & Timo Roettger"
format: html
editor: visual
execute:
  error: false
  warning: false
  message: false
callout-appearance: simple
---

{{< include 00-preamble.qmd >}}

# Exploring the data set

Let's first explore the data set which we will be using during the course. It is rich and fun.

We will be analyzing data from Experiment 1 of the following paper:

Scherbaum, S., & Kieslich, P. J. (2018). Stuck at the starting line: How the starting procedure influences mouse-tracking data. Behavior Research Methods, 50(5), 2097-2110.

[https://osf.io/7vrkz/](link%20to%20OSF)

It's a methodological paper but it replicates a very intuitive finding. When you are categorizing animals into types (e.g. a 'dog' into the type 'mammal'), you perform better when the animal is prototypical (dog \--\> mammal) as opposed to atypical (dolphin \--\> mammal). This 'performing better' is measured in different ways: reaction time, accuracy (correct vs. incorrect) and several measurement of the cursor movement toward the correct answer.

## Loading and inspecting the data

The data is part of the `aida` package, but we can give it a fancy new name:

```{r}
dolphin <- aida::data_MT
```

To get some information about the data set, we can use the `help` function:

```{r}
#| eval: false
help("data_MT")
```

Here is some more information we can get about the data:

```{r}

# number of rows in the data set
nrow(dolphin)

# number of columns in the data set
ncol(dolphin)

# names of the columns
names(dolphin)

# number of unique `subject_id`s
dolphin$subject_id |> unique() |> length()

# number of types each subject saw different `conditions`
dolphin |> with(table(subject_id, condition)) |> head()
```

``` {head(dolphin)}
```

## A closer look at the columns

Let's take a closer look at the columns and the information inside them.

We can get a `glimpse` of all columns like so:

```{r}
glimpse(dolphin)
```

Here is a quick explanation of all the different columns:

-   `trial_id` = unique id for individual trials
-   `MAD` = maximal deviation into competitor space
-   `AUC` = area under the curve
-   `xpos_flips` = the amount of horizontal direction changes
-   `RT` = reaction time in ms
-   `prototype_label` = different categories of prototypical movement strategies
-   `subject_id` = unique id for individual participants
-   `group` = groups differ in the response design (click vs. touch)
-   `condition` = category membership (Typical vs. Atypical)
-   `exemplar` = the concrete animal
-   `category_left` = the category displayed on the left
-   `category_right` = the category displayed on the right
-   `category_correct`= the category that is correct
-   `response` = the selected category
-   `correct` = whether or not the `response` matches `category_correct`

# Some basic data wrangling

Although the data set is already in good shape, here are some excursions into common steps to wrangle data (select, filter, reshape, summarize ...).

## Selecting columns

For now, we are only interested in columns `RT`, `group`, `condition`, `category_correct`, and `correct`. We can use the `select()` function of `dplyr` to get rid of columns we don't need.

```{r}

# selecting specific columns
dolphin_selected <-
  dolphin |>
  dplyr::select(RT, group, condition, category_correct, correct)
 
# let's have a look
dolphin_selected

```

## Filtering rows

If we care only about a subset of rows, we can use the `filter()` function. For example, let's filter all trials in which the correct category was either a `fish` or a `mammal`

```{r}

dolphin_filter1 <-
  dolphin_selected |> 
  filter(category_correct == "fish" | category_correct == "mammal")
  # the | is a logical operator that indicates that either the first expression OR 
  # the second one has to be true

dolphin_filter1

```

You can also `filter()` *against* particular conditions. For example, let's filter all rows that do not have `bird` as their correct category:

```{r}

dolphin_filter2 <-
  dolphin_selected |> 
  filter(category_correct != "bird")

dolphin_filter2

```

We can also filter according to multiple conditions at once, including numeric conditions. Here, we also filter for trials that have correct responses.

```{r}

dolphin_filter3 <-
  dolphin_selected |> 
  filter(category_correct != "bird",
         correct == 1)

dolphin_filter3

```

## Grouping and summarizing

We can also generate summary statistics of certain variables with a combination of `group_by()` & `summarise()`. Let's get the means and standard deviations of the reactions times for each level in the variable `condition`. We also include the minimum and maximum values for each condition.

```{r}

dolphin_aggregate <-
  dolphin_filter3 |>
  group_by(condition) |>
  summarise(
    min_RT  = min(RT),
    mean_RT = mean(RT, na.rm = T),
    sd_RT   = sd(RT, na.rm = T),
    max_RT  = max(RT)
    )
  # the na.rm = T is an argument that is used to tell R that NAs should be ignored 
  # when calculating the summary statistics

# show the aggregated df
dolphin_aggregate

```

So we find that atypical categories are responded to slower than typical categories. Makes sense. Identifying a dolphin as a mammal might be difficult because it shares a lot of features with fish.

We can group according to many different factors simultaneously, and we can create multiple summary statistics at the same time. Here, we get summary statistics for each combination of all levels in variables `condition` and `group`. We use the `tidyboot` package to get bootstrapped 95% confidence intervalls. (Notice that these are more informative than standard deviations in the sense that they give an upper and lower deviation, not just one number for both directions, which can be misleading when the data is skewed (like reaction times typically are)):

```{r aggregate2, warning = F, message=F}

dolphin_aggregate2 <-
  dolphin_filter3 |>
  group_by(group, condition) |>
  summarize(
    lower_CI = tidyboot::ci_lower(RT),
    mean_RT  = mean(RT, na.rm = T),
    upper_CI = tidyboot::ci_upper(RT)
    )

# show the aggregated df
dolphin_aggregate2

```

We can see here that the group that needed to click on a response are overall slower than touch responses, but also much more variable in their behavior.

## Changing and adding columns

Often, we are interested in standardized measures because we do not know what a value of 1 means on any given scale. Is 1 a large difference or a small difference? For example when we want to explore the impact of several predictors on the same measurement, we want to know the relative size of a number. To achieve this, we standardize measures by dividing their mean by their respective standard deviations. We will use the `scale()` function for this and create a new variable in our data frame via `mutate()`.

(Note that the `scale()` function creates an object that is of the matrix class. That is fine for the most part but might create issues later on. To avoid any issues, we wrap the `scale()` function in `as.numeric()` to store the results as a numeric vector.)

```{r standardize, warning = F, message=F}

dolphin_standardize <-
  dolphin_selected |>
  mutate(RT_scale = as.numeric(scale(RT)))
  
head(dolphin_standardize)

```

If we now compare, say atypical and typical categories according to reaction times, we can use the standardized RT ratings. Let's do all of this in one "pipeline".

```{r pipeline, warning = F, message=F}

dolphin_agg_standardize <- dolphin_selected |>
  mutate(RT_scale = scale(RT)) |> 
  group_by(condition) |>
  summarise(mean_RT_scale = mean(RT_scale, na.rm = T))
  
head(dolphin_agg_standardize)

```

Now we can see that atypical categories exhibit relatively higher RTs, i.e., more than 0.3 standard deviations higher than for typical categories.

## Exercises for data wrangling

### Exercise 1

::: callout-caution
**Exercise 1a**

Take the dolphin data set and store a reduced variant of it as `dolphin_reduced`. The new data frame should contain only the following columns: `RT`, `AUC`, `group`, and `exemplar`.
:::

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Show solution"

dolphin_reduced <- dolphin |>
  select(RT, AUC, group, exemplar)

head(dolphin_reduced)

```

::: callout-caution
**Exercise 1b**

We are for now only interested in those data that have whales as the exemplar. `filter()` only those rows and store them in a new dataframe called `whales_only`.
:::

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Show solution"

whales_only <- dolphin_reduced |> 
  filter(exemplar == "whale")

head(whales_only)

```

::: callout-caution
**Exercise 1c**

Now filter for only those data that have RTs below 1500ms.
:::

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Show solution"


whales_only2 <- whales_only |> 
  filter(RT < 1500)

head(whales_only2)

```

::: callout-caution
**Exercise 1d**

We don't like that `AUC` is unstandardized. Use `mutate()` to create a new vector that represents scaled AUC values (scaling is achieved by the function `scale()`).
:::

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Show solution"

whales_only_scaled <- whales_only2 |> 
  mutate(AUC_scaled = scale(AUC))

head(whales_only_scaled)

```

::: callout-caution
**Exercise 1e**

Calculate the mean scaled AUC ratings for both both groups.
:::

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Show solution"

whales_aggregate <- whales_only_scaled |> 
  group_by(group) |> 
  summarise(mean_AUC_scaled = mean(AUC_scaled, na.rm =TRUE))

head(whales_aggregate)

```

::: callout-caution
**Exercise 1f**

Do all of the above (a-e) in one pipeline.
:::

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Show solution"

whales_aggregate <- dolphin |>
  select(RT, AUC, group, exemplar) |> 
  filter(exemplar == "whale",
         RT < 1500) |> 
  mutate(AUC_scaled = scale(AUC)) |> 
  group_by(group) |> 
  summarise(mean_AUC_scaled = mean(AUC_scaled, na.rm =TRUE))
  
head(whales_aggregate)

```

### Exercise 2

::: callout-caution
**Exercise 1a**

Take the dolphin data set and store a reduced variant of it. The new data frame should contain only the columns `condition`, `group`, and `xpos_flips`, `correct`. And within the `correct` vector, we are only interested in the correct trials (= 1). Filter accordingly.
:::

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Show solution"

dolphin_sub <- dolphin |> 
  select(condition, group, xpos_flips, correct) |> 
  filter(correct == 1)

head(dolphin_sub)

```

::: callout-caution
**Exercise 2b**

Create an aggregated data frame that contains the mean `xpos_flips` value and the standard deviation for `group` and `condition`.
:::

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Show solution"

dolphin_agg <- dolphin_sub |>
  group_by(group, condition) |> 
  summarise(mean_xpos_flips = mean(xpos_flips, na.rm = TRUE),
            sd_xpos_flips = sd(xpos_flips, na.rm = TRUE))

head(dolphin_agg)

```

::: callout-caution
**Exercise 2c**

Use the `rename()` function to rename the new vectors for the mean `xflips` and their standard deviation to `xflips_mean` and `xflips_sd`.
:::

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Show solution"

dolphin_agg2 <- dolphin_agg |>
  rename(xflips_mean = mean_xpos_flips,
         xflips_sd = sd_xpos_flips)

```

::: callout-caution
**Exercise 2d**

Do all of the above (a-c) in one pipeline.
:::

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Show solution"

dolphin |> 
  select(condition, group, xpos_flips, correct) |> 
  filter(correct == 1) |> 
  group_by(group, condition) |> 
  summarise(mean_xpos_flips = mean(xpos_flips, na.rm = TRUE),
            sd_xpos_flips = sd(xpos_flips, na.rm = TRUE)) |> 
  rename(xflips_mean = mean_xpos_flips,
         xflips_sd = sd_xpos_flips)

```

# Basic plotting with `ggplot2`

To exercise with basic plotting functions, we first pick out a suitable subset of columns to work with:

```{r subset}

dolphin_subset <- dolphin |>
  dplyr::select(group, condition, RT, AUC, exemplar)

```

In order to plot summary statistics of our data (such as means and standard deviations), we need to create a data object with aggregated values.

```{r aggregate1}

# first group by 'condition' and then calculate the mean and standard deviation (sd) of `RT`
dolphin_agg <- dolphin_subset |>
  group_by(condition) |>
  summarise(
    lower_CI = tidyboot::ci_lower(RT),
    mean_RT = mean(RT),
    upper_CI = tidyboot::ci_upper(RT)
    )

dolphin_agg

```

## Basic plots

Now that we have pre-processed our data set, we are ready to visually explore it. Let's start very simple. Let's plot a bar plot. Let's also add a title to our plot.

```{r barebar, fig.width = 7, fig.align="center"}

ggplot(dolphin_agg, aes(x = condition, y = mean_RT)) +
  geom_bar(stat = "identity") +
  ggtitle("a bare bar plot")
  # stat = "identity" takes the number in the dataset as the bar height (as opposed to a 'count')

```

Ugh! What an ugly plot, right? But it's already telling a story: Atypical categories are responded to slower than typical categories. Let's add a measure of uncertainty, in our case the bootstrapped 95% confidence intervals, as error bars to the plot:

```{r}

ggplot(dolphin_agg, aes(x = condition, y = mean_RT)) +
  geom_bar(stat = "identity") + 
  
  # this is the added layer
  geom_errorbar(aes(ymin = lower_CI, 
                    ymax = upper_CI), 
                colour = "black",
                linewidth = 0.5) +
  
  ggtitle("a bare bar plot with error bars")
    
```

We can observe a couple of things here. First, ggplot automatically adjust the axes based on the elements to be plotted unless we tell it not to. Second, the error bars are plotted *in front* of the bars, i.e. closer to the viewer. This visual ordering reflects the order of layers. We first plotted the bars and THEN the error bars.

Beyond bar plots, we can create other useful plots types. For example a point plot. Instead of a bar, we plot the mean RT as points.

```{r point1, fig.width = 7, fig.align="center"}

ggplot(dolphin_agg, aes(x = condition, y = mean_RT)) +
  geom_errorbar(aes(ymin = lower_CI, 
                    ymax = upper_CI), 
                colour = "black") +  
  # this is the new geom 
  geom_point() +
  ggtitle("a point plot")

```

Or a line plot that connects the means with a line. For the line plot to work, we need to indicate a `group` aesthetic, i.e. the group that we want to connect with a line. If you have for example several interacting categories, you need to indicate which groups are supposed to be connected with lines (see below). Because we have only one group here, condition, we set `group` to `1`.

```{r line1, fig.width = 7, fig.align="center"}

ggplot(dolphin_agg, aes(x = condition, y = mean_RT, group = 1)) +
  geom_line() +
  ggtitle("a line plot")
 
```

Yay, we are on a roll. Let's plot a box plot. Remember the box shows the median (middle vertical line) and the interquartile range (the middle 50% of the data within the box). Note that for the box plot, we do not plot aggregated values, so we need to refer to the *entire* data set. We also add the aesthetic `fill` here and set it to the variable `condition` to color code our boxes.

```{r box1, fig.width = 7, fig.align="center"}

# we changed the dataset referred to
ggplot(dolphin, aes(x = condition, y = RT, fill = condition)) +
  # this is the new geom 
  geom_boxplot() +
  ggtitle("a box plot")

```

While the above plots illustrate one continuous variable (`RT`) plotted against a categorical variable (`condition`), we can also plot two continuous variables against each other. For example, we could plot `RT` against `AUC` in a scatter plot.

```{r scatter1, warning = F, fig.width = 6, fig.align="center"}

# we changed the y aesthetic to `Hardness`
ggplot(dolphin_subset, aes(x = RT, y = AUC)) +
  geom_point() +
  ggtitle("a scatter plot")

```

Finally, one central plot type for our class. The density plot. It plots on the x-axis a continous value and on the y-axis the "density"" of these values. So high values on the y-axis means a lot of data at the corresponding x-values. The density curve can be outlined with `color` and filled with `fill`. To keep the two categories visually distinct, we add an argument to the `geom_density()` function: alpha. Alpha controls the transparency of the color! We will see a lot of these density plots in our class.

```{r density1, warning = F, fig.width = 6, fig.align="center"}

ggplot(dolphin, aes(x = RT, color = condition, fill = condition)) +
  geom_density(alpha = 0.5) +
  ggtitle("a density plot")

```

## Adjusting plot elements

Okay, so we are now already capable of exploring our data visually with a bunch of plots. These plots are exceptionally ugly and of limited communicative value so far. Note that this is perfectly fine during an exploratory phase of data analysis. If we just eye-ball data and we have no trouble interpreting the plots, thats just fine. However, as soon as we want to communicate patterns to others with these graphs, we need to take a little bit more care of its communicative value. Let's look at ways we can tailor our plots to the needs of an audience.

Let's go back to our bar plot and explore whether `condition` and `group` has an impact on `RT`?

```{r bar3, fig.width = 4, fig.align="center"}

# First we aggregate RT for group and condition
dolphin_agg2 <- dolphin_subset |>
  group_by(group, condition) |>
  summarise(mean_RT = mean(RT),
            sd_RT = sd(RT))

# then we plot and color code for condition (note that, for bar plots, the aesthetic of `color` refers to the border of the bar, and `fill` refers to the actual colour of the bar)

ggplot(dolphin_agg2, aes(x = condition, y = mean_RT, fill = group)) +
  geom_bar(stat = "identity")

```

Hm... that doesn't work out, the bars are on top of each other, we need to tell ggplot to position the bars next to each other instead. We do that with position_dogde(). Note that ggplot assigns a default color coding scheme to your plots if you don't specify it by hand.

```{r bar4, fig.width = 6, fig.align="center"}

ggplot(dolphin_agg2, aes(x = condition, y = mean_RT, fill = group)) +
  geom_bar(stat = "identity", position = position_dodge())

```

Awww much better! Alternatively, we can plot the two categories into separate panels. We achieve this by facetting using the `facet_grid()` function.

```{r bar5, fig.width = 6, fig.align="center"}

ggplot(dolphin_agg2, aes(x = group, y = mean_RT, fill = condition)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  # this is the facetting function
  facet_grid(~ condition)

```

Okay. We are getting somewhere. We already learned something again. Apparently, the effect of typiciality on RT is pretty similar across tasks. It does not look like we have an interaction here (more on interactions later).

Now let's make these plots ready to communicate information. We add appropriate axes titles. Note that we are using a little hack here: The "\n" inserts an empty line, creating visual distance from axis title to axis, thus making it easier to read it. Our audience will thank us.

```{r bar6, fig.width = 6, fig.align="center"}

ggplot(dolphin_agg2, aes(x = group, y = mean_RT, fill = condition)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  facet_grid(~ condition) +
  # add axis titles
  xlab("\n task") +
    ylab("mean response latency in ms\n") 

```

The same graph as a point / line plot indicated to the audience whether there is an interaction pattern or not. Note that here we do not facet because we actually want the points to be plotted within the same horizontal space. We also have to specify the group aesthetic to tell ggplot which points to connect with lines.

```{r line2, fig.width = 8, fig.align="center"}

ggplot(dolphin_agg2, aes(x = group, y = mean_RT, color = condition, group = condition)) +
  # instead of geom_bar we use geom_point and geom_line
  geom_point(size = 12) +
  geom_line(size = 2) +
  xlab("\n task") +
  ylab("mean response latency in ms\n") 
  # # need to change to the color aesthetic instead of fill
  # scale_y_continuous(expand = c(0, 0), breaks = (c(0, 500, 1000, 1500, 2000, 2500, 3000)), limits = c(0,3000))

```

These lines look pretty parallel and don't indicate a strong interaction pattern. But how do different exemplars differ? Let's aggregate for individual exemplars first and then create the same plot for the means of all exemplars.

```{r line3, fig.width = 8, fig.align="center"}

dolphin_agg3 <- dolphin_subset |> 
  group_by(exemplar, group, condition) |> 
  summarise(mean_RT = mean(RT, na.rm = TRUE))

ggplot(dolphin_agg3, aes(x = group, y = mean_RT, color = condition, group = exemplar)) +
  # instead of geom_bar we use geom_point and geom_line
  geom_point(size = 6, alpha = 0.3) +
  geom_line() +
  geom_label(aes(label = exemplar)) +
  xlab("\n task") +
  ylab("mean response latency in ms\n")

```

It looks like "shark" and "rattlesnake" behave very different from their buddies in the `typical` condition. Interesting! We wouldn't have noticed if we had only looked at the overall means.

## Exercises for plotting

Take the scatter plot below as a departure point. It plots AUC (area-under-the-curve) against MAD (maximal absolute deviation).

```{r}

ggplot(dolphin, aes(x = MAD, y = AUC)) +
  geom_point() +
  ggtitle("a scatter plot")

```

::: callout-caution
**Exercise 3a**

(1) Change both the x-axis and the y-axis title to sensible and informative titles.

(2) Change the plot title to something informative.

(3) Change the scaling of the x-axis to display only MAD values between -500 and 500
:::

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Show solution"

ggplot(dolphin, aes(x = MAD, y = AUC, 
                           
                           # (d) add color aesthetic
                           color = group)) +
  geom_point() +
  
  # (1) axes titles
  xlab("\n maximal absolute deviation") +
  ylab("area-under-the-curve \n") +
  
  # (2) change title
  ggtitle("MAD is correlated with AUC") +

  # (3) change x-axis (note that certain values are not displayed then. R will spit out a warning)
  scale_x_continuous(limits = c(-500,500))

```

::: callout-caution
**Exercise 4**

(1) Plot AUC values as a function of group in a density plot (`geom_density`).

(2)) Make the density curves semi-transparent with the `alpha` argument

(3)) Add the aida_theme to the plot.

(4) Add the mean values for both groups into the density plot as a line.
:::

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Show solutions for a-c"

# (1 - 3)
ggplot(dolphin, aes(x = AUC, color = group, fill = group)) +
  geom_density(alpha = 0.3) +
  xlab("\n AUC")

```

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Show solution for d"

# (4) aggregate means and add to plot

dolphin_agg <- dolphin |>
  group_by(group) |>
  summarise(mean_AUC = mean(AUC, na.rm = TRUE),
            sd_AUC = sd(AUC, na.rm = TRUE))

# add them to the plot as vertical lines
ggplot(dolphin, aes(x = AUC, color = group, fill = group)) +
  geom_density(alpha = 0.3) +
  xlab("\n AUC") +
  # since the vertical line refers to dolphin_agg, we need to specify the dataset explicitly 
  geom_vline(data = dolphin_agg, 
             aes(xintercept = mean_AUC, color = group),
             lty = "dashed") 
 
```
