---
title: "02c: Categorical predictors (exercises)"
subtitle: "Bayesian regression: theory & practice"
author: "Michael Franke & Timo Roettger"
format: html
editor: visual
execute:
  error: false
  warning: false
  message: false
callout-appearance: simple
---

# Preamble

{{< include 00-preamble.qmd >}}

```{r}
dolphin <- aida::data_MT
```

# Regression w/ multiple categorical predictors

We want to regress `log RT` against the full combination of categorical factors `group`, `condition`, and `prototype_label`.

`log RT ~ group * condition * prototype_label`

The research hypotheses we would like to investigate are:

1.  Typical trials are faster than atypical ones.
2.  CoM trials are slower than the other kinds of trials (straight and curved) together, and respectively.
3.  'straight' trials are faster than 'curved' trials.
4.  Click trials are slower than touch trials.

But for this to work (without at least mildly informative priors), we would need to have a sufficient amount of observations in each cell. So, let's check:

```{r}
dolphin |>
  mutate(group = as_factor(group),
         condition = as_factor(condition),
         prototype_label = as_factor(prototype_label)) |>
  count(group, condition, prototype_label, .drop = FALSE) |>
  arrange(n)
```

So, there are cells for which we have no observations at all. For simplicity, we therefore just lump all "change of mind"-type trajectories into one category:

```{r}
dolphin_prepped <-
  dolphin |>
  mutate(
    prototype_label = case_when(
     prototype_label %in% c('curved', 'straight') ~ prototype_label,
     TRUE ~ 'CoM'
    ),
    prototype_label = factor(prototype_label,
                             levels = c('straight', 'curved', 'CoM')))

dolphin_prepped |>
  select(RT, prototype_label)
```

Here is a plot of the data to be analyzed:

```{r}
dolphin_prepped |>
  ggplot(aes(x = log(RT), fill = condition)) +
  geom_density(alpha = 0.4) +
  facet_grid(group ~ prototype_label)

```

## Exercise 1

::: callout-caution
**Exercise 1a**

Use `brm()` to run a linear regression model for the data set `dolphin_prepped` and the formula:

`log RT ~ group * condition * prototype_label`

Set the prior for all population-level slope coefficients to a reasonable, weakly-informative but unbiased prior.
:::

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Show solution"

fit <- brm(
  formula = log(RT) ~ group * condition * prototype_label,
  prior   = prior(student_t(1, 0, 3), class = "b"),
  data    = dolphin_prepped
  )
```

::: callout-caution
**Exercise 1b**

Plot the posteriors for population-level slope coefficients using the `tidybayes` package in order to:

1.  determine which combination of factor levels is the default cell
2.  check which coefficients have 95% CIs that do *not* include zero
3.  try to use this latter information to address any of our research hypotheses (stated above)
:::

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Show solution"
tidybayes::summarise_draws(fit)
tidybayes::gather_draws(fit, `b_.*`, regex = TRUE) |>
  filter(.variable != "b_Intercept") |>
  ggplot(aes(y = .variable, x = .value)) +
  tidybayes::stat_halfeye() +
  labs(x = "", y = "") +
  geom_segment(x = 0, xend = 0, y = Inf, yend = -Inf,
               lty = "dashed")

# the default cell is for click-atypical-straight

# coeffiencents with 95% CIs that do not include zero are:
#   grouptouch, conditionTypical, prototype_labelCoM

# none of these give us direct information about our research hypotheses
```

::: callout-caution
**Exercise 1c**

Use the `faintr` package to get information relevant for the current research hypotheses. Interpret each result with respect to what we may conclude from it.
:::

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Show solution"

# 1. Typical trials are faster than atypical ones.
# -> There is overwhelming evidence that this is true
#    (given the data and the model).

faintr::compare_groups(
  fit,
  lower  = condition == 'Typical',
  higher = condition == 'Atypical'
)

# 2. CoM trials are slower than the other kinds of trials
#    (straight and curved) together, and respectively.
# -> There is overwhelming evidence that this is true
#    (given the data and the model).

faintr::compare_groups(
  fit,
  lower  = prototype_label != 'CoM',
  higher = prototype_label == 'CoM'
)

faintr::compare_groups(
  fit,
  lower  = prototype_label == 'straight',
  higher = prototype_label == 'CoM'
)

# 3. 'straight' trials are faster than 'curved' trials.
# -> There is no evidence for this hypothesis
#    (given the data and the model).

faintr::compare_groups(
  fit,
  lower  = prototype_label == 'straight',
  higher = prototype_label == 'curved'
)

# 4. Click trials are slower than touch trials.
# -> There is overwhelming evidence that this is true
#    (given the data and the model).

faintr::compare_groups(
  fit,
  lower  = group == 'touch',
  higher = group == 'click'
)
```

# Regression w/ metric & categorical predictors

# Exercise 2

::: callout-caution
**Exercise 2a**

Create a new dataframe that contains only the mean values of the RT, and MAD for each animal (`exemplar`) and for correct and incorrect responses. Print out the `head` of the new dataframe.
:::

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Show solution"

# aggregate
dolphin_agg <- dolphin |> 
  group_by(exemplar, correct) |> 
  dplyr::summarize(MAD = mean(MAD, na.rm = TRUE),
                   RT = mean(RT, na.rm = TRUE))
  
# let's have a look
head(dolphin_agg)

```

::: callout-caution
**Exercise 2b**

Run a linear regression using brms. `MAD` is the dependent variable (i.e. the measure) and both `RT` and `correct` are independent variables (`MAD ~ RT + correct`). Tip: the coefficients might be really really small, so make sure the output is printed with enough numbers after the comma.
:::

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Show solution"

# specify the model 
model2 = brm(
  # model formula
  MAD ~ RT + correct, 
  # data
  data = dolphin_agg
  )

print(summary(model2), digits = 5)

```

Try to understand the coefficient table. There is one coefficient for `RT` and one coefficient for `correct` which gives you the change in MAD from incorrect to correct responses.

::: callout-caution
**Exercise 2c**

Plot a scatter plot of MAD \~ RT and color code it for correct responses (Tip: Make sure that `correct` is treated as a factor and not a numeric vector). Draw two predicted lines into the scatterplot. One for correct responses ("lightblue") and one for incorrect responses ("orange").
:::

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Show solution"

dolphin_agg$correct <- as.factor(as.character(dolphin_agg$correct))

# extract model parameters:
model_intercept <- summary(model2)$fixed[1,1]
model_RT <- summary(model2)$fixed[2,1]
model_correct <- summary(model2)$fixed[3,1]

# plot
ggplot(data = dolphin_agg, 
       aes(x = RT, 
           y = MAD,
           color = correct)) + 
  geom_abline(intercept = model_intercept, slope = model_RT, color = "orange", size  = 2) +
  geom_abline(intercept = model_intercept + model_correct , slope = model_RT, color = "lightblue",size  = 2) +
  geom_point(size = 3, alpha = 0.3)

```

::: callout-caution
**Exercise 2d**

Extract the posteriors for the coefficients of both `RT` and `correct` from the model output (use the `spread_draws()` function), calculate their means and a 67% Credible Interval. Print out the `head` of the aggregated dataframe.
:::

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Show solution"

# get posteriors for the relevant coefficients
posteriors2 <- model2 |>
  spread_draws(b_RT, b_correct) |>
  select(b_RT, b_correct) |> 
  gather(key = "parameter", value = "posterior")

# aggregate
posteriors2_agg <- posteriors2 |> 
  group_by(parameter) |> 
  summarise(mean_posterior = mean(posterior),
            `67lowerCrI` = HDInterval::hdi(posterior, credMass = 0.67)[1],
            `67higherCrI` = HDInterval::hdi(posterior, credMass = 0.67)[2]
            )

# print out
posteriors2_agg
  
```

::: callout-caution
**Exercise 2e**

Plot the scatterplot from 2c and plot 50 sample tuples for the regression lines for correct and incorrect responses.
:::

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Show solution"

# sample 50 random numbers from 4000 samples
random_50 <- sample(1:4000, 50, replace = FALSE)
  
# wrangle data frame
posteriors3 <- model2 |>
  spread_draws(b_Intercept, b_RT, b_correct) |>
  select(b_Intercept, b_RT, b_correct) |> 
  # filter by the row numbers in random_50
  slice(random_50)
  
# plot
ggplot(data = dolphin_agg, 
       aes(x = RT, 
           y = MAD, 
           color = correct)) + 
  geom_abline(data = posteriors3,
              aes(intercept = b_Intercept, slope = b_RT), 
              color = "orange", size  = 0.1) +
  geom_abline(data = posteriors3,
              aes(intercept = b_Intercept + b_correct, slope = b_RT), 
              color = "lightblue", size  = 0.1) +
  geom_point(size = 3, alpha = 0.3)

```

::: callout-caution
**Exercise 2f**

Given our model and our data, calculate the evidential ratio of correct responses exhibiting larger MADs than incorrect responses.
:::

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Show solution"

hypothesis(model2, 'correct > 0')

```

# Exercise 3

Here is an aggregated data set `dolphin_agg` for you.

```{r exercise1}

# aggregate
dolphin_agg <- dolphin %>% 
  group_by(group, exemplar) %>% 
  dplyr::summarize(MAD = median(MAD, na.rm = TRUE),
                   RT = median(RT, na.rm = TRUE)) %>% 
  mutate(log_RT = log(RT))

```

::: callout-caution
**Exercise 3a**

Standardize ("z-transform") `log_RT` such that the mean is at zero and 1 unit corresponds to the standard deviation. Name it `log_RT_s`.
:::

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Show solution"

dolphin_agg$log_RT_s <- scale(dolphin_agg$log_RT, scale = TRUE)

```

::: callout-caution
**Exercise 3b**

Run a linear model with `brms` that predicts `MAD` based on `log_RT_s`, `group`, and their two-way interaction.
:::

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Show solution"

model1 = brm(
  MAD ~ log_RT_s * group, 
  data = dolphin_agg
  )

```

::: callout-caution
**Exercise 3c**

Plot `MAD` (y) against `log_RT_s` (x) in a scatter plot and color-code for `group`. Plot the regression lines for the click and the touch group into the plot and don't forget to take possible interactions into account.
:::

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Show solution"

# extract posterior means for model coefficients
Intercept = summary(model1)$fixed[1,1]
log_RT = summary(model1)$fixed[2,1]
group = summary(model1)$fixed[3,1]
interaction = summary(model1)$fixed[4,1]

# plot
ggplot(data = dolphin_agg, 
       aes(x = log_RT_s, 
           y = MAD, 
           color = group)) + 
  geom_point(size = 3, alpha = 0.3) +
  geom_vline(xintercept = 0, lty = "dashed") +
  geom_abline(intercept = Intercept, slope = log_RT, 
              color = project_colors[1], size = 2) +
  geom_abline(intercept = Intercept + group, slope = log_RT + interaction, 
              color = project_colors[2], size = 2) 

```

::: callout-caution
**Exercise 3d**

Specify very skeptic priors for all three coefficients. Use a normal distribution with mean = 0, and sd = 10. Rerun the model with those priors.
:::

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Show solution"

# specify priors
priors_model2 <- c(
  set_prior("normal(0,10)", class = "b", coef = "log_RT_s"),
  set_prior("normal(0,10)", class = "b", coef = "grouptouch"),
  set_prior("normal(0,10)", class = "b", coef = "log_RT_s:grouptouch")
)

# model
model2 = brm(
  MAD ~ log_RT_s * group, 
  data = dolphin_agg,
  prior = priors_model2
  )

```

::: callout-caution
**Exercise 3e**

Compare the model output of model1 to model2. What are the differences and what are the reasons for these differences?
:::

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Show solution"

# We can compare the model predictions by looking at the coefficients / plotting them:
summary(model1)
summary(model2)

# extract posterior means for model coefficients
Intercept = summary(model2)$fixed[1,1]
log_RT = summary(model2)$fixed[2,1]
group = summary(model2)$fixed[3,1]
interaction = summary(model2)$fixed[4,1]

# plot
ggplot(data = dolphin_agg, 
       aes(x = log_RT_s, 
           y = MAD, 
           color = group)) + 
  geom_point(size = 3, alpha = 0.3) +
  geom_vline(xintercept = 0, lty = "dashed") +
  geom_abline(intercept = Intercept, slope = log_RT, 
              color = project_colors[1], size = 2) +
  geom_abline(intercept = Intercept + group, slope = log_RT + interaction, 
              color = project_colors[1], size = 2) 

# ANSWER: the magnitude of the coefficients is much smaller in model2, with the interaction term being close to zero. As a result, the lines in the plot are closer together and run in parallel. The reason for this change lies in the priors. We defined the priors of model2 rather narrowly, down weighing  data points larger or smaller than zero. This is a case of the prior dominating the posterior.

```
