---
title: "Stochastic independence"
subtitle: "Bayesian regression: theory & practice"
author: "Michael Franke"
format: html
execute:
  error: false
  warning: false
  message: false
  cache: true
callout-appearance: simple
editor:
  markdown:
    wrap: sentence
---

# Preamble

{{< include 00-preamble.qmd >}}

# Using regression modeling to test for stochastic independence

Here are three data sets, each with triples of associated observations for variables $X$, $Y$, and $Z$.
Ideally, you do not look at the code, so you will have more fun with the exercise below.

```{r}

##################################################
# creating three data sets with various dependencies
##################################################

data_1 <- tibble(
  X = rnorm(500),
  Z = map_dbl(X, function(i) rnorm(1, mean = i)),
  Y = map_dbl(Z, function(i) rnorm(1, mean = i))
)

data_2 <- tibble(
  Z = rnorm(500),
  X = map_dbl(Z, function(i) rnorm(1, mean = i)),
  Y = map_dbl(Z, function(i) rnorm(1, mean = i))
)

data_3 <- tibble(
  X = rnorm(500),
  Y = rnorm(500),
  Z = map2_dbl(X, Y, function(i,j) rnorm(1, mean = i+j)),
)

```


::: {.callout-caution collapse="false"}
## Exercise 1: Use regression to assess stochastic independence

Use linear regression to investigate whether:

a. $X$ and $Y$ are stochastically independent
b. $X$ and $Y$ are stochastically independent given Z

in each of these data sets.

Ideally, write a single function that takes a data set as input and output information that tells you about whether $X$ and $Y$ are independent.

::: {.callout-tip collapse="true"}
### Solution

Here is a reusable function:

```{r}
#| results: hide

test_stochDep <- function(input_data) {
  
  fit_simple   <- brms::brm(Y ~ X, data = input_data)
  fit_multiple <- brms::brm(Y ~ X + Z, data = input_data)
  
  result_simple <- brms::hypothesis(fit_simple, "X>0")
  result_multiple <-brms::hypothesis(fit_multiple, "X>0")
  
  return(list(simple = result_simple, multiple = result_multiple))
}

result_1 = test_stochDep(data_1)
result_2 = test_stochDep(data_2)
result_3 = test_stochDep(data_3)

```

In data set 1, $X$ and $Y$ are not stochastically independent, but turn independent given $Z$.

```{r}
result_1
```

In data set 2, $X$ and $Y$ are not stochastically independent, and turn independent given $Z$.

```{r}
result_2
```

In data set 3, $X$ and $Y$ are stochastically independent, but cease to be given $Z$.

```{r}
result_3
```

:::
:::

::: {.callout-caution collapse="false"}
## Exercise 2: Interpret your results

Infer which (if any) of the three data sets was created with one of the following causal structures:

- collider: $X, Y \rightarrow Z$
- chain:    $X \rightarrow Z \rightarrow Y$
- fork:     $Z \rightarrow X,Y$

::: {.callout-tip collapse="true"}
### Solution

Data sets 1 and 2 could be from a chain or a fork, because $X$ and $Y$ are not stochastically independent, but turn independent given $Z$.
We cannot say which based on these tests.

Data set 3 could be from a collider, because $X$ and $Y$ are stochastically independent, but cease to be given $Z$.


:::
:::

